---
title: "merge options"
author: "o.duke-williams@ucl.ac.uk"
date: "03/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a short RMarkdown file showing how merge() works with inputs where the common field between two tables has some missing values.

Firstly, let's create two sample data frames, df1 and df2

```{r create-data}
df1 <- data.frame(id=c("a","b","d"),v1=c("red","blue","green"))
df2 <- data.frame(id=c("a","b","c","e"),v2=c("up","down","left","right"))
``` 

## Tasks

You will be asked to add some elements to this file. You will need to open it in RStudio. For each task you will need to add some R code. You will then need to save the RMarkdown file and then use 'knit'. If there are problems in your code, you will see error messages in the windos.

### Task

First, add some code to display the content of both data frames

```{r task1}
# You will add code here
summary(df1)
summary(df2)
```

We are now going to merge the data frames, and look at the results

```{r merge1}
m1 <- merge(df1,df2)
m1
```

How many rows are shown in the results? Why do you think that these rows are included.

### Task

Edit the RMarkdown file to complete the sentence:

The merged table m1 shows 3 rows. This is because it has to match the number of rows as one of the df's.

We can adjustthe behaviour of merge using the 'all' option. Using 'all.x=TRUE' we indicate that we will use all the values in the 'x' table (the first one mentioned in the merge() statement, in our case df1).

```{r merge2}
m2 <- merge(df1,df2,all.x=TRUE)
m2
```

### Task

Edit the RMarkdown file to complete the sentence:

The merged table m2 shows *num* rows. This is because...

What value do the results show where df1 contained an id that was not found in df2?

### Task

Edit the R code in the markdown file to do a merge of df1 and df2 using all.y=TRUE. Make sure you display the result.

What is the effect this time?

```{r task2}
# You will add code here
m3 <- merge(df1,df2,all.y=TRUE)
m3
```

### Task

Finally, edit the R code in the markdown file to do a merge of df1 and df2 using all=TRUE.

What is the effect this time? How does this differ from all.x and all.y?

```{r task3}
# You will add code here
m4 <- merge(df1,df2,all=TRUE)
m4
```
